---
title: "DBTree"
author: "Rutger Vos (@rvosa)"
date: "29-8-2019"
output:
  pdf_document: default
  html_document: default
---

This document demonstrates the usage of DBTree databases in R. Whereas many
other languages tend to make access to relational databases more palatable 
by providing object-relational mappings (possibly generated from the schema),
the more common approach in R is simply to present the contents of a database
table as a data frame. This is what we will do here, so let's set things up.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RSQLite)
library(dplyr)
library(dbplyr)
library(ape)
```

We are going to query the plant phylogeny of 
[Smith & Brown, 2018](https://doi.org/10.1002/ajb2.1019). Release version
v1.0 of this tree has been preprocessed with `megatree-loader` and deposited
on figshare, which we will now download into a `data` subfolder:

```{r download}
dir.create("data", showWarnings = FALSE)
download.file(
    url = "https://ndownloader.figshare.com/files/17461988",
    destfile = "data/ALLMB.db", 
    mode = "wb")
```

The database just contains a single (relevant) table. We will now connect to
the database and instantiate that table (called `node`) as if it was a data
frame:

```{r connect}
ALLMB.db <- DBI::dbConnect(RSQLite::SQLite(), "data/ALLMB.db")
ALLMB.tre <- tbl(ALLMB.db, "node")
```

Let's attempt to create a patristic distance matrix for the members of genus
[Abelia](https://www.ncbi.nlm.nih.gov/taxonomy/86866). We start by querying the 
database for all tips whose name matches the substring:

```{r tips}
tips <- ALLMB.tre %>%
    filter(name %like% 'Abelia%') %>%
    arrange(left) %>%
    collect()
```

For all pairs of tips we are now going to look up the MRCA of the pair, and then
compute the distance between the tips. I'm sure there are more efficient ways to
do this but it shows how we can use the schema:

```{r dist}
ntips <- nrow(tips)
x <- c()
for ( i in 1:ntips-1 ) {
    for ( j in i+1:ntips ) {
    
        # get the MRCA for the pairwise comparison
        minl <- min(tips[i,]$left,tips[j,]$left)
        maxr <- max(tips[i,]$right,tips[j,]$right)
        mrca <- ALLMB.tre %>%
            filter(left < minl && right > maxr) %>%
            arrange(desc(left)) %>%
            head(1) %>%
            collect()
        
        # calculate the distance
        mrcah <- mrca$height
        pdist <- (tips[i,]$height-mrcah)+(tips[j,]$height-mrcah)
        x <- c(x, pdist)
    }
}
```

The distances we now have form a lower triangle. As per the instructions in 
ape::nj, we convert these to a matrix, perform neighbor-joining, and plot the 
result:

```{r matrix}
M <- matrix(0, ntips, ntips)
M[lower.tri(M)] <- x
M <- t(M)
M[lower.tri(M)] <- x
dimnames(M) <- list(tips$name)
tr <- nj(M)
plot(tr, "u")
```

